---
title: 'Titanic: Kaggle'
author: "Sandra Ezidiegwu"
date: "August 26, 2016"
output: html_document
---
# Introduction
This is my first attempt at a kaggle dataset. Decided to give it a try to test my data exploratory abilities. Feel free to share comments too... Enjoy!

### Import useful libraries
```{r message=FALSE, warning=FALSE}
library(rpart) #classification algorithm
library(rpart.plot) #visualization
library(ggplot2) #visualization
library(dplyr) #data manipulation
setwd("/Users/sandraezidiegwu/Documents/Data Science/Titanic/")
```

## Data Collection and Gathering
```{r}
genderclass <- read.csv("genderclassmodel.csv", header = T, sep = ",")
gender <- read.csv("gendermodel.csv", header = T, sep = ",")
test <- read.csv("test.csv", header = T, sep = ",")
train <- read.csv("train.csv", header = T, sep = ",")
```

## Date Exploration
```{r}
names(test)
names(train)
test$Survived <- 0
all.pass <- rbind(train, test)
str(all.pass)
summary(all.pass)
```

VARIABLE NAME | DESCRIPTIONS
--------------|--------------
Survived      |(0 = No; 1 = Yes)
Pclass        |Passenger Class (1 = 1st; 2 = 2nd; 3 = 3rd)
Name          |Passenger Name
Sex           |Passenger's Sex
Age           |Passenger's Age
SibSp         |Number of Siblings/Spouses Aboard
Parch         |Number of Parents/Children Aboard
Ticket        |Ticket Number
Fare          |Passenger Fare
Cabin         |Cabin
Embarked      |Port of Embarkation (C = Cherbourg; Q = Queenstown; S = Southampton)

Let's dig deeper into columns that might affect survival rate. Eg. Title, Family Size, Age etc.
```{r}
all.pass$Title <- gsub('(.*, )|(\\..*)', '', all.pass$Name)
table(all.pass$Sex, all.pass$Title)

all.pass$Title[all.pass$Title %in% c('Mlle', 'Ms', 'Lady')] <- 'Miss'
all.pass$Title[all.pass$Title %in% c('Dona', 'the Countess', 'Capt', 'Col', 'Don','Jonkheer', 'Major')] <- 'Affluent'
all.pass$Title[all.pass$Title %in% c('Dr', 'Master', 'Rev', 'Sir')] <- 'Mr'
all.pass$Title[all.pass$Title %in% 'Mme'] <- 'Mrs'
table(all.pass$Sex, all.pass$Title)

ggplot(all.pass, aes(x = Title, fill = factor(Survived))) +
geom_bar(stat = 'count', position = 'dodge') 
```

Create Family Size column
```{r}
all.pass$FSize <- all.pass$SibSp + all.pass$Parch + 1
```

Aggregate by family name
```{r}
all.pass$FName <- paste(all.pass$Surname, all.pass$FSize, sep = '_')
```

Vizualize the relationship between family size and survival
```{r}
ggplot(all.pass, aes(x = FSize, fill = factor(Survived))) + 
  geom_bar(stat = 'count', position = 'dodge') +
  scale_x_continuous(breaks = c(1:11)) +
  labs(x = 'Family Size') +
  theme_linedraw()
```

The barplot shows that the larger the family size, the lower the chances of survival. I'm also assuming that affluent families had a lower family size count. Lets check that... 
```{r}
ggplot(all.pass, aes(x = FSize, fill = factor(Title))) + 
  geom_bar(stat = 'count', position = 'dodge') +
  scale_x_continuous(breaks = c(1:11)) +
  labs(x = 'Family Size') +
  theme_linedraw()
```

I was right! The largest family size for affluent families is 3 so they had a good chance of survival. Let's vsualize this further...

```{r}
par(mfrow = c(1,1))
mosaicplot(table(all.pass$Title, all.pass$Survived), main = 'Survival by Title', shade = TRUE)
```
You can see that the 'affluent' folks pretty much survived.. Not sure about the Capt. though, according to the movie, he sank with the ship.. But hey! What do I know...

## Missing Values
```{r}
sum(is.na(all.pass$Fare))
na.fare <- all.pass[is.na(all.pass$Fare),]
na.fare
```

PassengerId 1044 has an NA Fare value. He/She ranks in 3rd class and embarked form 'S' however.We will replace row 153 fare value with the median fare value of that class and embarkment.
```{r}
all.pass$Fare[153] <- median(all.pass[all.pass$Pclass == '3' & all.pass$Embarked == 'S',]$Fare, na.rm = T)

par(mfrow = c(1,1))
hist(all.pass$Age, main = 'Age Distribution aboard the Titanic', xlab = "Age", col = 'lightblue')
sum(is.na(all.pass$Age))

actual <- all.pass
actual2 <- all.pass
age.part <- rpart(Age ~ Pclass+Sex+SibSp+Parch+Ticket+Fare+Cabin+FSize+Title, data = all.pass[!is.na(all.pass$Age),], method = 'anova')
age.pred <- predict(age.part, all.pass[is.na(all.pass$Age),])
actual$Age[is.na(actual$Age)] <- age.pred
```

For accuracy, let's prune some variables to achieve a better prediction
```{r}
age.prune <- prune.rpart(age.part, cp = 0.1)
age.pred2 <- predict(age.prune, all.pass[is.na(all.pass$Age),])
actual2$Age[is.na(actual2$Age)] <- age.pred2

mean(actual2$Age != actual$Age)
```
There's a 20% improvement on error with pruned prediction

Visualize age distributions of the actual data and the predicted values
```{r}
par(mfrow = c(1,3))
hist(all.pass$Age, col = 'blue', main = 'Actual Age Values', xlab = 'Age')
hist(actual$Age, col = 'lightblue', main = 'Predicted Age Values', xlab = 'Age')
hist(actual2$Age, col = 'lightblue', main = 'Pruned Predicted Age Values', xlab = 'Age')
```

Pruned prediction data is more accurate so let's go ahead and replace the missing age values with the pruned prediction age values and check for na's
```{r}
all.pass$Age[is.na(all.pass$Age)] <- age.pred2
sum(is.na(all.pass$Age))
all.pass$AgeDist[all.pass$Age < 18] <- "Child"
all.pass$AgeDist[all.pass$Age >= 18] <- "Adult"
all.pass$AgeDist <- factor(all.pass$AgeDist)
```

Let's evaluate survival count by age
```{r}
table(all.pass$AgeDist, all.pass$Survived)
```

## Prediction
Let's move on to predict the survival. We start by creating test and training data sets
```{r}
test_data <- all.pass[892:1309,]
train_data <- all.pass[1:891,]
test_data$Survived <- NULL
```

We are ready to build our model using random Forest!
```{r}
model_part <- rpart(Survived ~ Pclass + Sex + Age + SibSp +
                             Parch + Fare + Title + FSize + AgeDist, data = train_data, method = 'class', control = rpart.control(cp=0.0001))
```

Visualize decision tree
```{r}
par(mfrow=c(1,1))
rpart.plot(model_part)
```

Let's predict the test data set and view our solution
```{r}
model_pred <- predict(model_part, test_data, type = 'class')

solution <- data.frame(PassengerID = test_data$PassengerId, Survived = model_pred)
solution[1:10,]
```

## Solution 
Write solution to csv file
```{r}
write.csv(solution, file = 'rpart-solution.csv', row.names = F)
```

View Solution
```{r}
final_solution <- read.csv('rpart-solution.csv', sep = ',', header = TRUE)
final_solution[1:10,]
```
